Avoidance in permutations

### 2014-11-07 project2/paper/paper.tex
some edits to section 6 + todo

\begin{theorem}
The number of SYT of shape $(n, n)$ is given by $C_n$, the $n^{th}$ Catalan number.
\end{theorem}

@@

\begin{theorem}
There exists an algorithm (called the \emph{RSK algorithm}) which defines a bijection between the set of permutations in $S(n)$ avoiding a decreasing subsequence of length 3 and the set of pairs of size-$n$ SYT of the same shape and at most two rows. 
\end{theorem}\todo{give a citation for this}
Using all this machinery and all these results, we are finally able to prove the closed-form expression for $s_n(321)$.
\begin{theorem}
The total number of permutations of $\{1,2,3,\ldots, n\}$ that avoid the order 321 as a subsequence is $C_n$, where $C_n$ is the $n^{th}$ Catalan number.
\end{theorem}
\begin{proof}
By the first part of Theorem 12, there are $s_n(321)$ elements in the set $A$ of pairs of size-$n$ SYT of the same shape and at most two rows. By Theorem 11, the set $B_n$ of SYT of shape $(n, n)$ has its size given by $C_n$. We will construct a bijection between the sets $A_n$ and $B_n$: because $|A_n|=s_n(321)$ and $|B_n|=C_n$, it will therefore follow that $s_n(321)=C_n$, as desired.

@@

\begin{theorem}
$t_n(312)$, the set of permutations in $T_n$ avoiding 312, is precisely the same size as $t_n(321)$, the set of permutations in $T_n$ avoiding 321.
\end{theorem}

@@

Now if $\pi$ does not avoid $321$, there must exist some sub-sequence $x, y, z$ in $\pi$ in the relative ordering $321$. From Lemma \ref{swapping_lemma}, we see that swapping $y$ and $z$ in the permutation $\pi$ yields another valid permutation $\pi_1$ in $T_n$ -- note that the permutation produced after this swap has $x, y, z$ in the relative ordering $312$ instead of $321$. Note that $\pi_1$ may still have sub-sequences in the relative ordering $321$, so we may have to perform the above swapping operation again to get some permutation $\pi_2 \in T_n$ such that a sub-sequence $x_1, y_1, z_1$ in $\pi_1$ that has a relative ordering $321$ is in the relative ordering $312$ in $\pi_2$. Continuing this process produces the following chain of permutations,
$$\pi \rightarrow \pi_1 \rightarrow \pi_2 \rightarrow \ldots \rightarrow \pi_k$$
Where the last permutation in the chain $\pi_k$ does not have a single three-element sub-sequence in the relative ordering $321$, that is $\pi' = \pi_k$ is a permutation in $T_n$ that avoids $321$.

@@

The above two claims shows there's actually an easy way to transform an arbitrary permutation $a \in A$ to a $b \in B$, and similarly, to transform an arbitrary permutation $b \in B$ to a $a \in A$ -- to obtain a permutation in $B$ from a permutation $a \in A$, for every element $i$ in the permutation $a$, all even elements in $a$ to the right of $i$ less than $i$ should be reversed (that is reordered to in increasing order); similarly to obtain a permutation in $A$ from a permutation $b \in B$, for every element $i$ in the permutation $b$, all even elements in $b$ to the right of $i$, less than $i$ should be reversed (that is reordered to in decreasing order). Since the mapping from $A$ to $B$ is clearly reversible, and hence invertible, we conclude that the mapping we just established between $A$ and $B$ is bijective.

>>>>

\begin{theorem}
\label{catalan_2}
The number of SYT of shape $(n, n)$ is given by $C_n$, the $n^{th}$ Catalan number.
\end{theorem}

@@

\begin{theorem}
\label{rsk}
There exists an algorithm (called the \emph{RSK algorithm}) which defines a bijection between the set of permutations in $S(n)$ avoiding a decreasing subsequence of length 3 and the set of pairs of size-$n$ SYT of the same shape and at most two rows. 
\end{theorem}\todo{give a citation for this}
Using all this machinery and all these results, we are finally able to prove the closed-form expression for $s_n(321)$.
\begin{theorem}
The total number of permutations of $\{1,2,3,\ldots, n\}$ that avoid the order 321 as a subsequence is $C_n$, where $C_n$ is the $n^{th}$ Catalan number.
\end{theorem}
\begin{proof}
By Theorem \ref{rsk}, there are $s_n(321)$ elements in the set $A$ of pairs of size-$n$ SYT of the same shape and at most two rows. By Theorem \ref{catalan_2}, the set $B_n$ of SYT of shape $(n, n)$ has its size given by $C_n$. We will construct a bijection between the sets $A_n$ and $B_n$: because $|A_n|=s_n(321)$ and $|B_n|=C_n$, it will therefore follow that $s_n(321)=C_n$, as desired.

@@

\begin{theorem}
There exists a bijection between the set of permutations in $T_n$ avoiding 312 and the set of permutations in $T_n$ avoiding 321. As a consequence, $t_n(312)=t_n(321)$.
\end{theorem}

@@

Now if $\pi$ does not avoid $321$, there must exist some sub-sequence $x, y, z$ in $\pi$ in the relative ordering $321$. From Lemma \ref{swapping_lemma}, we see that swapping $y$ and $z$ in the permutation $\pi$ yields another valid permutation $\pi_1$ in $T_n$ -- note that the permutation produced after this swap has $x, y, z$ in the relative ordering $312$ instead of $321$. Note that $\pi_1$ may still have sub-sequences in the relative ordering $321$, so we may have to perform the above swapping operation again to get some permutation $\pi_2 \in T_n$ such that a sub-sequence $x_1, y_1, z_1$ in $\pi_1$ that has a relative ordering $321$ is in the relative ordering $312$ in $\pi_2$. Continuing this process produces the following chain of permutations,
$$\pi \rightarrow \pi_1 \rightarrow \pi_2 \rightarrow \ldots \rightarrow \pi_k$$
Where the last permutation in the chain $\pi_k$ does not have a single three-element sub-sequence in the relative ordering $321$, that is $\pi' = \pi_k$ is a permutation in $T_n$ that avoids $321$.\todo{why does this terminate? also, why is this necessary?}

@@

The above two claims shows there's actually an easy way to transform an arbitrary permutation $a \in A$ to a $b \in B$, and similarly, to transform an arbitrary permutation $b \in B$ to a $a \in A$ -- to obtain a permutation in $B$ from a permutation $a \in A$, for every element $i$ in the permutation $a$, all even elements in $a$ to the right of $i$ less than $i$ should be reversed (that is reordered to in increasing order); similarly to obtain a permutation in $A$ from a permutation $b \in B$, for every element $i$ in the permutation $b$, all even elements in $b$ to the right of $i$, less than $i$ should be reversed (that is reordered to in decreasing order). \todo{should have some justification here that reversing reordering these elements doesn't 'step on each other'... that is, you can always do it for some $i_1$ and then some $i_2$ afterward}
Since the mapping from $A$ to $B$ is clearly reversible, and hence invertible, we conclude that the mapping we just established between $A$ and $B$ is bijective.
